Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2019-08-30 paper/results.tex
moving comments to comments

% word limit: 500
\section{Results}
\label{sec:results}

@@

\begin{figure}
  \caption{
      Top: Rotation period vs effective temperature for stars in the \mct\
    sample, colored by the velocity dispersions of stars over a grid in
    $\log_{10}$(period) and \teff.
    Bottom: the velocity dispersions of groups of stars, shown as a solid
    grid for clarity.
    The black solid lines on both panels show a 1.1 Gyr isochrone, calculated
    with the \citet{angus2019} gyrochronology relation which roughly traces
    the rotation period gap.
    The black dashed lines show a 650 Myr isochrone, indicating the location
    and shape of the Praesepe cluster (to which this gyrochronology model was
    calibrated).
    The black points show the 1.1 Gyr NGC 6811 open cluster.
}
  \centering
    \includegraphics[width=1\textwidth]{vplot}
\label{fig:dispersion_period_teff}
\end{figure}
Figure \ref{fig:dispersion_period_teff} shows rotation period vs. \teff\ for
our sample, coloured by (\vb) velocity dispersion.
Velocity dispersion was calculated for groups of stars over a grid in
$\log_{10}$(period) and temperature.
If we assume that mass dependent heating doesn't affect this sample, and \vb\
at low galactic latitudes is an unbiased tracer of \vz, \vb\ velocity
dispersion can be interpreted as an age proxy and stars plotted in a similar
color in figure \ref{fig:dispersion_period_teff} are similar ages.
Interpreted this way, lines of constant age (isochrones) appear to follow the
shape of the Praesepe-based gyrochronology model (black dotted line) at young
ages.
However, this does not seem to be true age old ages where it appears that
rotation period {\it decreases} with decreasing effective temperatures.
This is a paradigm shift for gyrochronology because stellar spin-down rate is
thought to be directly tied to magnetic field strength, and the deeper
convection zones of cooler stars generate stronger magnetic fields which {\it
should} lead to more efficient angular momentum loss.
% The shape of the upper envelope of rotation periods in the top panels of
% figures \ref{fig:age_cut} and \ref{fig:dispersion_period_teff} also hints at
% this.
The shape of the period-\teff\ relations at old ages appears to follow the
shape of the upper detection edge, with the so-called `M dwarf dip'
\citep{vansaders2018} reflected in the lines of constant velocity dispersion
(and presumed age) in the top panel of figure
\ref{fig:dispersion_period_teff}.
If the shape of that upper edge is created by a detection limit, our results
suggest that stars cooler than $\sim$4500 K become inactive (and their
rotation periods therefore become undetectable) at around the same age, but at
an older age than stars hotter than $\sim$4500.
The (kinematically) oldest stars in the \mct\ sample are cooler than 4500 K.
This is probably because cooler stars remain active for longer, and their
rotation periods can therefore be measured at older ages.

We reiterate that the velocity dispersions of the coolest stars may be
affected by a selection bias -- these extremely faint stars are more difficult
to detect at larger distances, larger heights above the plane, and therefore
larger velocities.
It is possible that some high velocity stars of temperatures cooler than
$\sim$4000 K are missing from this sample, and the velocity dispersions may
therefore appear lower than they truly are.
This could be why the velocity dispersion appears to decrease towards the
right of figure \ref{fig:dispersion_period_teff}.

>>>>

% word limit: 500
\section{Results and discussion}
\label{sec:results}

@@

\begin{figure}
  \caption{
      Top: Rotation period vs effective temperature for stars in the \mct\
    sample, colored by the velocity dispersions of stars over a grid in
    $\log_{10}$(period) and \teff.
    Bottom: the velocity dispersions of groups of stars, shown as a solid
    grid for clarity.
    The black solid lines on both panels show a 1.1 Gyr isochrone, calculated
    with the \citet{angus2019} gyrochronology relation which roughly traces
    the rotation period gap.
    The black dashed lines show a 650 Myr isochrone, indicating the location
    and shape of the Praesepe cluster (to which this gyrochronology model was
    calibrated).
    The black points show the 1.1 Gyr NGC 6811 open cluster.
}
  \centering
    \includegraphics[width=1\textwidth]{vplot}
\label{fig:dispersion_period_teff}
\end{figure}
Figure \ref{fig:dispersion_period_teff} shows rotation period vs. \teff\ for
our sample, coloured by (\vb) velocity dispersion.
Velocity dispersion was calculated for groups of stars over a grid in
$\log_{10}$(period) and temperature.
If we assume that mass dependent heating does not strongly affect this sample,
and \vb\ at low galactic latitudes is an unbiased tracer of \vz, \vb\ velocity
dispersion can be interpreted as an age proxy and stars plotted in a similar
color in figure \ref{fig:dispersion_period_teff} are similar ages.
Interpreted this way, lines of constant age (isochrones) appear to follow the
shape of the Praesepe-based gyrochronology model (black solid and dotted
lines) at young ages.
However, at old ages it appears that rotation period {\it decreases} with
decreasing effective temperatures.
This is a paradigm shift for gyrochronology because stellar spin-down rate is
thought to be directly tied to magnetic field strength, and the deeper
convection zones of cooler stars generate stronger magnetic fields which {\it
should} lead to more efficient angular momentum loss.
The shape of the period-\teff\ relations at old ages appears to follow the
shape of the upper detection edge, with the so-called `M dwarf dip'
\citep{vansaders2018} reflected in the lines of constant velocity dispersion
(and presumed age) in the top panel of figure
\ref{fig:dispersion_period_teff}.
If the shape of that upper edge is created by a detection limit, our results
suggest that stars cooler than $\sim$4500 K become inactive (and their
rotation periods therefore become undetectable) at around the same age, but at
an older age than stars hotter than $\sim$4500.
The (kinematically) oldest stars in the \mct\ sample are cooler than 4500 K.
This is presumably because cooler stars remain active for longer, and their
rotation periods can therefore be measured at older ages.

The velocity dispersions of the coolest stars may be affected by a selection
bias -- these extremely faint stars are more difficult to detect at larger
distances, larger heights above the plane, and therefore larger velocities.
It is possible that some high velocity stars of temperatures cooler than
$\sim$4000 K are missing from this sample, and the velocity dispersions may
therefore appear lower than they truly are.
This could be why the velocity dispersion appears to decrease towards the
right of figure \ref{fig:dispersion_period_teff}.

\subsection{The period gap}
\label{sec:period_gap}

The origin of the rotation period gap, first identified
by \citet{mcquillan2013} and visible in figures \ref{fig:age_cut} and
\ref{fig:dispersion_period_teff} still remains a mystery.
This gap can be seen as an under-density of points between the 0.7-1.0 and
1.0-1.5 Gyr age ranges in figure \ref{fig:age_cut} and roughly follows a line
of constant gyrochronal age of around 1.1 Gyr \citep[according to the
gyrochronology relation of][]{angus2019}, as shown in figure
\ref{fig:dispersion_period_teff}.
Several explanations for the gap's origin have been proposed, including a
discontinuous star formation history \citep{mcquillan2013, davenport2017,
davenport2018}, a rapid change in magnetic field structure
\citep{reinhold2019}, and erroneous rotation period measurements that are
incorrect by a factor of two \citep{koen2018}.
The latter explanation can be ruled out because stars below the gap have
smaller velocity dispersions than the stars above the gap, indicating that
they are kinematically younger \citep{mcquillan2013, davenport2018}, as
evident in figures \ref{fig:age_cut} and \ref{fig:dispersion_period_teff}.
For stars below the gap, in the 0.7-1.0 Gyr age range shown in figure
\ref{fig:age_cut}, velocity dispersion is relatively constant as a function of
temperature, however above the gap, in the 1.0-1.5 Gyr age range and older,
velocity dispersion increases with \teff.
The coolest stars in the 1.0-1.5 Gyr age range have the same velocity
dispersion as the hottest stars in the age range above which indicates that
the period-\teff\ relations are {\it flat} at these rotation periods.
This is also visible in figure \ref{fig:dispersion_period_teff}.
Below the gap, velocity dispersion within a given period range appears to {\it
decrease} with decreasing temperature.
The opposite appears to be true above the gap.

In the $\sim$ 1.1 Gyr NGC 6811 cluster, the rotation periods of mid-K dwarfs
are faster than expected; their rotational evolution appears to have stalled,
and the period-\teff\ relation is flat \citep{curtis2019}.
The rotation periods of the K dwarfs in this cluster are plotted in figure
\ref{fig:dispersion_period_teff}.
Although the G dwarfs in this cluster fall on the 1.1 Gyr gyrochronology
model, the K dwarfs lie only a little above the 0.65 Gyr gyrochronology model.
NGC 6811 straddles the rotation period gap: its G dwarfs lie above it and its
K dwarfs lie below it.
This cluster may be the `missing link' that connects two epoch of stellar
spin-down: an early stage where the period-\teff\ relation for K dwarfs has a
negative slope and a late stage where it has a positive slope.
In this case the period gap may delineate the transition between these two
regimes and is the point at which stellar magnetic dynamos likely undergo a
dramatic structural shift at an age of $\sim$ 1.1 Gyr.

It may be a coincidence that the gyrochronology relations seem to only flatten
off {\it above} the period gap, or it may be that the period gap is located at
magnetic transition period.
We lack a sufficient quantity of data to investigate in detail but new field
star rotation periods from the \ktwo\ and \tess\ missions may be able to
validate or rule out this hypothesis in the future.
